---
title: "Automatic Detection"
author: "Howard Baek"
date: "6/29/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(factoextra)
library(raster)
library(tmap)
library(png)
library(gridExtra)
library(dendextend)
library(DescTools)
library(imageryML)
library(plotly)
library(ggmap)
theme_set(theme_light())
```

## Notes from paper

Upwelling is defined by a certain temperature drop compared to the surrounding water temperature close to the coast. Such strong upwelling events could contribute to replenishing the euphotic zone with the nutritional components necessary for biological productivity.

Automatic Detection Method: A simple temperature threshold value was specified. Upwelling was detected by calculating the temperature difference for each individual pixel from the zonal mean temperature, for every pixel line. 

Two different temperature thresholds: 2 degrees / 3.5 degrees.


## Data Preprocessing: Daily data

Latitude (42.625, 52.125) and Longitude (229.875, 236.625) 

### Degree Conversion:

Longitude: 230 degree East - 360 = -130 degree West

```{r}
# I downloaded the data using read_csv("https://coastwatch.pfeg.noaa.gov/erddap/griddap/ncdcOisst21Agg.csv?sst%5B(2010-01-01T12:00:00Z):(2021-06-01T12:00:00Z)%5D%5B(0.0)%5D%5B(42.625):(52.125)%5D%5B(229.875):(236.625)%5D&.draw=surface&.vars=longitude%7Clatitude%7Csst&.colorBar=%7C%7C%7C12%7C19%7C&.bgColor=0xffccccff")
# This big data is saved as pilot_data.csv

# Raw data (last 10 years)
df_raw <- read_csv("howard-analysis/pilot_data.csv")

# Processed data
df_processed <- df_raw %>% 
  # Get rid of miscellaneous zlev in first row
  slice(-1) %>% 
  # zlev is a column of zeroes, so get rid of that
  dplyr::select(-zlev) %>% 
  # Convert into date
  mutate(time = ymd_hms(time)) %>% 
  # Set column names
  rename(date = time,
         lat = latitude,
         lon = longitude) %>% 
  # Convert date column to Date type
  mutate(date = as.Date(date),
         lat = as.numeric(lat),
         lon = as.numeric(lon),
         sst = as.numeric(sst))

# mask out Puget Sound, Strait of Juan de Fuca and Georgia Strait
masks <- list(c(235.4488, 236.884, 47.87651, 50.13138),
              c(232.2913, 233.8987, 50.28689, 51.60871),
              c(234.4154, 235.9654, 49.04283, 50.09251))

for (m1 in masks) {
  # index of Puget Sound, Strait of Juan de Fuca or Georgia Strait
  mask_loc <- df_processed$lat <= m1[4] & df_processed$lat >= m1[3] &
    df_processed$lon <= m1[2] & df_processed$lon >= m1[1]
  # Change to NA
  df_processed$sst[mask_loc] <- NA
}
```

Have a look at raw data on plot
```{r}
df_processed %>% 
  filter(date == "2010-01-01") %>% 
  ggplot() +
  geom_point(aes(lon, lat)) +
  geom_point(aes(lon, lat), data = df_processed %>% 
               filter(date == "2010-01-01",
                      is.nan(sst) |
                      is.na(sst)),
             color = "red") +
  geom_text(aes(lon, lat,label = sst),
            data = df_processed %>% 
              filter(date == "2010-01-01") %>% 
              sample_n(500),
            vjust = 0.5, hjust = 1) +
  geom_vline(xintercept = 236.625, color = "midnightblue") +
  geom_hline(yintercept = 52.125, color = "midnightblue") 
```


Stamen map with SST
```{r}
# Create sample data (2010-01-01)
sample_df <- df_processed %>% 
  filter(date == "2010-01-01") %>% 
  mutate(lon = lon - 360)

# Give coordinates for map
bbox <- c(left = 229.875 - 360, bottom = 42.625, right = 236.625 - 360, top = 52.125)
# Get stamen map
ocean_map <- get_stamenmap(bbox, zoom = 8)

# Plot
ggmap(ocean_map) +
  geom_tile(aes(lon, lat, fill = sst),
            alpha = 0.4,
            data = sample_df) +
  geom_rect(xmin = -125, xmax = -124.2,
            ymin = 43, ymax = 44.4,
            fill = NA,
            color = "red",
            size = 1) +
  scale_fill_gradient2(midpoint = mean(sample_df$sst, na.rm = TRUE),
                       low = "blue",
                       mid = "white",
                       high = "red") +
  labs(x = NULL,
       y = NULL,
       fill = "Sea Surface Temperature (SST)",
       title = "Latitude (42.625, 52.125) and Longitude (229.875, 236.625)")
```


Algorithm to detect SST at coast (border between ocean and land)

- 28 longitude values for each latitude
- 2 degree threshold

```{r}
# Find which latitude has the most NA or NaN SST values
sample_df %>% 
  group_by(lat) %>% 
  filter(is.na(sst) | is.nan(sst)) %>% 
  count(lat, sort = TRUE)

# Find3 SST values next to land (sst_coast_X)
is_upwelling <- sample_df %>% 
  group_by(lat) %>% 
  # 1 longitude tick away from land
  mutate(sst_coast_1 = last(na.omit(sst)),
         #2 longitude ticks away from land
         sst_coast_2 = nth(na.omit(sst), -2),
         # 3 longitude ticks away from land
         sst_coast_3 = nth(na.omit(sst), -3),
         # 4 longitude ticks away from land
         sst_coast_4 = nth(na.omit(sst), -4)) %>% 
  # Find difference between pixels
  # Threshold: 0.05
  summarize(is_upwelling_1_2 = abs(sst_coast_1 - sst_coast_2) > 0.15,
            is_upwelling_2_3 = abs(sst_coast_2 - sst_coast_3) > 0.15,
            is_upwelling_3_4 = abs(sst_coast_3 - sst_coast_4) > 0.15) %>% 
  ungroup() %>% 
  group_by(lat) %>% 
  # Check if any upwelling in each latitude
  # first() because I want to return one row for each lat
  summarise(is_upwelling_total = first(is_upwelling_1_2 | is_upwelling_2_3 | is_upwelling_3_4)) %>% 
  ungroup()

# upwelling latitudes
is_upwelling
```


- Latitude & Longitude: every 0.25 degrees


Function to detect upwelling and plot on stamen map

- Join `sample_data` and `is_upwelling` to get `is_upweling_total` column into sample_data
- Find (coast) longitude corresponding to row preceding last Na/NaN value in SST

```{r}
final_df <- sample_df %>% 
  left_join(is_upwelling) %>%
  ungroup() %>% 
  group_by(lat) %>% 
  # Find  (coast) longitude corresponding to row preceding last Na/NaN value in SST
  mutate(last_lon = across(sst, ~ tail(lon[!is.na(.)], 1))) %>% 
  ungroup() %>% 
  mutate(last_lon = last_lon$sst)


# Create df for rectangle coordinates
rect_coord <- final_df %>% 
  filter(is_upwelling_total) %>% 
    select(lat, last_lon) %>% 
    unique() %>% 
    mutate(xmin = last_lon - 0.1,
           xmax = last_lon + 0.1,
           ymin = lat - 0.1,
           ymax = lat + 0.1) %>% 
  rename(xmin_test = xmin,
         xmax_test = xmax,
         ymin_test = ymin,
         ymax_test = ymax)
```


```{r}
# Plot
ggplot() +
  geom_tile(aes(lon, lat, fill = sst),
            alpha = 0.4,
            data = final_df) +
  geom_rect(data = rect_coord,
            mapping = aes(xmin = xmin_test, xmax = xmax_test,
                          ymin = ymin_test, ymax = ymax_test),
            size = 1,
            fill = NA,
            color = "red") +
  scale_fill_gradient2(midpoint = mean(final_df$sst, na.rm = TRUE),
                       low = "blue",
                       mid = "white",
                       high = "red") +
  labs(x = NULL,
       y = NULL,
       fill = "Sea Surface Temperature (SST)",
       title = "Latitude (42.625, 52.125) and Longitude (229.875, 236.625)")
```